Computing likelihoods under the SMC for a general class of ARGs.

### 2023-12-13 paper.tex
fix references

\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argneelde}[0]{\texttt{ARG-Needle}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}
\newcommand{\tskit}[0]{\texttt{tskit}}

@@

For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history where all events are uniquely positioned 
both in time and along the genome [Wong et al. 2023].
Yet, the ease by which such a fully detailed history can be simulated under the 
CwR, is not matched by the usability of the model for inference. 
In particular, given an ARG, there will always be an infinite amount of other 
ARGs that could have generated the observed genomic dataset 
[potentially with very different likelihoods under the CwR]
\citep{mcvean_2005_approximating}.

@@

To improve the ability to simulate the CwR, McVean and Cardin 
provided an approximation of the CwR that is both Markovian backwards in time 
as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
By sacrificing precision on our ability to capture long range linkage 
information, the SMC became the engine of many powerful inference methods [PSMC]. 
A later improvement, the SMC', closed the gap between the SMC and the 
CwR even further.

The second key approximation is not a generative model. Rather than providing 
a mathematical description of the genealogies underlying the observed data 
the Li and Stephens or copying model pragmatically details the relationship 
of a single genome to a larger set of genomes as the result of an 
incomplete copying process \citep{li_2003_modeling}. This heuristic approach 
has been successfully applied for demographic inference [and other things: ADD] 
\citep{sheehan_2013_estimating, steinrucken_2018}.

Both approaches have further formed the basis (of many) of the recent advancements in 
ARG inference methods [Rasmussen et al. 2014, Heine et al. 2018, 
Kelleher et al. 2019, Speidel et al. 2019,
Rasmussen et al. 2022, Zhang et al.2023]. 
A further classification of these methods could be made based on the level of detail 
that is provided on recombination events. The first real breakthrough in large-scale 
ARG inference, \argweaver, heads the first group. 
The ARGs sampled by \argweaver from the posterior distribution 
given the observed variational data, are all fully precise and detailed in the sense 
that all events are explicitly inferred. Note however that the set of possible node times 
has to be limited to enable the usage of an HMM. The other more recent methods, although 
still SMC-based, have added heuristic pre- and/or post-processing steps 
that have allowed them to extend their approach to larger sample sizes. We should also 
mention \kwarg and \arginfer in this category of methods producing fully detailed ARGs 
although these are not based on the SMC or the copying algorithm. In fact they represent 
two extremes along the approximation continuum with \kwarg being entirely 
parsimony-based while \arginfer is a CwR-based MCMC sampler.

@@

The computational gains from not having to pinpoint recombination events tie in 
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS]. 
Some approaches have embraced the approximate nature of the resulting ARG by, for 
example, highlighting the inability to resolve certain splits and instead representing 
them as polytomies \citep{kelleher_2019_inferring}. Taken together this means that the 
degree of completeness in which genetic inheritance from ancestors to descendants is documented 
by each of these methods varies extensively.
[TO ADD: these methods only provide point estimates]

@@

ARG simplification is thus linked to the presence of unary nodes in 
the local trees and simultaneously touches upon the question of what is 
knowable. Many ARG reconstruction methods return a data structure that 
is in essence a series of local tries [REFs]. 
As highlighted, this does not need to be a contradiction. 
All ARGs can be represented as a series of local trees \citep{wong-2023}.
The important question is therefore is however about the extent by which any 
ARG reconstruction methods incorporates the available information 
on how nodes persist across these local trees.\\

@@

As a final remark, the described algorithm relies heavily on the presence of locally unary 
nodes in the tree sequence to correctly identify the full extent of all lineages going back 
through time. Although their presence is crucial to our ability embed any local tree in the 
ARG, the concept and their importance are still relatively new (but see 
\citet{Wong-2023}). Although we have briefly touched upon the observation that in 
the presence of unary nodes we seem to augment the left-to-right logic of the SMC 
with some additional information. 
We believe that the topic of locally unary nodes in general requires further exploration. 
And more specifically, we see it as future work to fully quantify to what extent the 
absence of these locally unary nodes will affect the shape of the likelihood surface given 
any set of parameters.\\

>>>>

\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argneedle}[0]{\texttt{ARG-Needle}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}
\newcommand{\tskit}[0]{\texttt{tskit}}

@@

For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history where all events are uniquely positioned 
both in time and along the genome [Wong et al. 2023].
Yet, the ease by which such a fully detailed history can be simulated under the 
CwR, is not matched by the usability of the model for inference. 
In particular, given an ARG, there will always be an infinite amount of other 
ARGs that could have generated the observed genomic dataset 
[potentially with very different likelihoods under the CwR]
\citep{mcvean_approximating_2005}.

@@

To improve the ability to simulate the CwR, McVean and Cardin 
provided an approximation of the CwR that is both Markovian backwards in time 
as well as left-to-right along the genome \citep{mcvean_approximating_2005}. 
By sacrificing precision on our ability to capture long range linkage 
information, the SMC became the engine of many powerful inference methods [PSMC]. 
A later improvement, the SMC', closed the gap between the SMC and the 
CwR even further.

The second key approximation is not a generative model. Rather than providing 
a mathematical description of the genealogies underlying the observed data 
the Li and Stephens or copying model pragmatically details the relationship 
of a single genome to a larger set of genomes as the result of an 
incomplete copying process \citep{li_modeling_2003}. This heuristic approach 
has been successfully applied for demographic inference [and other things: ADD] 
\citep{sheehan_estimating_2013, steinrucken_inference_2019}.

Both approaches have further formed the basis (of many) of the recent advancements in 
ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018, kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}. 
A further classification of these methods could be made based on the level of detail 
that is provided on recombination events. The first real breakthrough in large-scale 
ARG inference, \argweaver, heads the first group. 
The ARGs sampled by \argweaver from the posterior distribution 
given the observed variational data, are all fully precise and detailed in the sense 
that all events are explicitly inferred. Note however that the set of possible node times 
has to be limited to enable the usage of an HMM. The other more recent methods, although 
still SMC-based, have added heuristic pre- and/or post-processing steps 
that have allowed them to extend their approach to larger sample sizes. We should also 
mention \kwarg and \texttt{ARGinfer} in this category of methods producing fully detailed ARGs 
although these are not based on the SMC or the copying algorithm. In fact they represent 
two extremes along the approximation continuum with \kwarg being entirely 
parsimony-based while \texttt{ARGinfer} is a CwR-based MCMC sampler.

@@

The computational gains from not having to pinpoint recombination events tie in 
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS]. 
Some approaches have embraced the approximate nature of the resulting ARG by, for 
example, highlighting the inability to resolve certain splits and instead representing 
them as polytomies \citep{kelleher_inferring_2019}. Taken together this means that the 
degree of completeness in which genetic inheritance from ancestors to descendants is documented 
by each of these methods varies extensively.
[TO ADD: these methods only provide point estimates]

@@

ARG simplification is thus linked to the presence of unary nodes in 
the local trees and simultaneously touches upon the question of what is 
knowable. Many ARG reconstruction methods return a data structure that 
is in essence a series of local tries [REFs]. 
As highlighted, this does not need to be a contradiction. 
All ARGs can be represented as a series of local trees \citep{wong_general_2023}.
The important question is therefore is however about the extent by which any 
ARG reconstruction methods incorporates the available information 
on how nodes persist across these local trees.\\

@@

As a final remark, the described algorithm relies heavily on the presence of locally unary 
nodes in the tree sequence to correctly identify the full extent of all lineages going back 
through time. Although their presence is crucial to our ability embed any local tree in the 
ARG, the concept and their importance are still relatively new (but see 
\citet{wong_general_2023}). Although we have briefly touched upon the observation that in 
the presence of unary nodes we seem to augment the left-to-right logic of the SMC 
with some additional information. 
We believe that the topic of locally unary nodes in general requires further exploration. 
And more specifically, we see it as future work to fully quantify to what extent the 
absence of these locally unary nodes will affect the shape of the likelihood surface given 
any set of parameters.\\
